\shortdescription{Seeing different geometries as a variation on a theme.}
\activitytitle{Introduction}

As a young mathematician I was introduced to the classic \textit{Le\c{c}ons
sur la G\'{e}om\'{e}trie des Espaces de Riemann}, written by the great French
geometer \'{E}lie Cartan. Early in his treatise on geometries in all
dimensions, the author presents the case of two-dimensional geometries, in
particular, those two-dimensional geometries that look the same at all points
and in all directions. (For example, a cylinder looks the same at each of its
points but not in all directions emanating from any one of its points, whereas
a sphere looks the same at all points and in all directions.) It turns out
that there is one and only one such geometry for each real number $K$, called
the \textit{curvature} of the geometry. The case $K=0$ is the (flat) Euclidean
geometry that you learned in high school.

These twenty-five pages of Cartan's book (Chapter VI, \S i-v) so captivated me
that I have returned to them regularly throughout my career and have adapted
and taught them many times at the advanced undergraduate level. They form the
basis for this little book. To me they tell one of the most beautiful and
satisying stories in all of geometry, one which exemplifies a fundamental
principle of all great mathematics, namely that, using the tools at hand but
in a slightly novel way, the clouds part and one sees that objects and
relationships that seemed so different are in fact parts of a single elegant story!

When $K>0$ it turns out that the $K$-geometry is the geometry of the sphere of
radius $R=1/K^{1/2}$ that we can see as a subset of Euclidean three-space
$\mathbb{R}^{3}$. But the geometries with $K<$ $0$ are not so easy to
visualize. They are the so-called `hyperbolic' geometries. In fact it took
mathematicians a couple thousand years to realize that the existed at all! It
turns out that the secret to understanding all the two-dimensional geometries,
including the ones with $K<0$, in a unified way is to simply rescale the third
coordinate in $\mathbb{R}^{3}$ and, in these `unusual' coordinates $\left(
x,y,z\right)  $, to look at each two-dimensional geometry as the solution set
to the equation%
\[
K\left(  x^{2}+y^{2}\right)  +z^{2}=1.
\]
However the idea of changing coordinates without changing the underlying
geometry described by those coordinates is a challenging one that did not come
into mathematics until a couple of centuries ago. It will require that, before
we get into the beautiful uniform study of all two-dimensional geometries, we
practice the coordinate change we are going to use, namely the rescaling of
the third coordinate in Euclidean $3$-space. That practice, together with a
review of some concepts from several variable calculus and linear algebra,
will comprise Part \ref{I} of this book.

It has often been said that \textquotedblleft mathematics is not a spectator
sport." This truism is very much in evidence in the writing of this book. It
is written so as to guide you through the entire story, yet permit you, when
possible, to construct the mathematical story for yourself, that is, to do
some mathematics yourself rather than just observe it done by others. This
`doing mathematics oneself' takes the form of Exercises with enough help
(Hints) provided so that the `doing' is not so onerous as to get in the way of
the story itself.

Strong evidence has been provided by students of mathematics over many
centuries that such guided `doing' is indispensible for understanding and
retention. In fact the very form of this book, as a loose-leaf or electronic
notebook, is intended to encourage you to write out (in correctable form)
solutions to the Exercises that can be inserted at the appropriate places into
the text.

This book supposes familiarity with several variable calculus and the linear
algebra of matrices. In particular you will need to remember and apply the
chain rule for differentiable functions of several variables, written in
matrix notation. That is:

\begin{theorem}
\textbf{(Chain Rule)} Given differentiable mappings%
\begin{align*}
&  \left(  y_{1}\left(  x_{1},\ldots,x_{m}\right)  ,\ldots,y_{n}\left(
x_{1},\ldots,x_{m}\right)  \right) \\
&  \left(  z_{1}\left(  y_{1},\ldots,y_{n}\right)  ,\ldots,z_{p}\left(
y_{1},\ldots,y_{n}\right)  \right)
\end{align*}
then%
\[
\left(  \frac{\partial z_{k}}{\partial x_{i}}\right)  =\left(  \frac{\partial
z_{k}}{\partial y_{j}}\right)  \cdot\left(  \frac{\partial y_{j}}{\partial
x_{i}}\right)  .
\]

\end{theorem}

As a help, at some points in the text and in some of the Exercises, a more
complete treatment of a particular topic can be found in one of the following
two texts:

[MJG]: Greenberg, Marvin Jay. \textit{Euclidean and Non-Euclidean Geometry:
Development and History.} W.H. Freeman \& Co. 3rd Ed., 1994.

[DS]: Davis, H. and Snider, A.D. \textit{Introduction to Vector Analysis.} Wm.
C. Brown Publishers, 7th Ed. 1994.

The corresponding topics in these texts are referenced. For example, [MJG,311]
refers to page 311 in the Greenberg book and [DS,59ff] refers to page 59 and
those pages just following page 59 in the Davis-Snider book.

Some final remarks about notation in this book. The letters `\textbf{EG}' will
always mean Euclidean (usually plane but occasionally $3$-dimensional)
Geometry, the letters `\textbf{SG}' will always mean Spherical Geometry, and
the letters `\textbf{HG}' will always mean Hyperbolic Geometry. One further
kind of geometry, which we call Neutral Geometry, will be explained in the
book and denoted by `\textbf{NG}.'

Also, it will often be useful to consider a vector, for example, $V=\left(
a,b,c\right)  $, as a $1\times3$ matrix and we will write%
\[
\left(  V\right)  =\left(
\begin{array}
[c]{ccc}%
a & b & c
\end{array}
\right)
\]
or as a $3\times1$ matrix and we will write%
\[
\left(  V\right)  ^{t}=\left(
\begin{array}
[c]{c}%
a\\
b\\
c
\end{array}
\right)  .
\]
This will allow us, for example, to write the scalar product of two vectors%
\begin{align*}
V\bullet V^{\prime}  &  =\left(  a,b,c\right)  \bullet\left(  a^{\prime
},b^{\prime},c^{\prime}\right) \\
&  =aa^{\prime}+bb^{\prime}+cc^{\prime}%
\end{align*}
as a product of matrices%
\[
V\cdot V^{\prime}=\left(
\begin{array}
[c]{ccc}%
a & b & c
\end{array}
\right)  \cdot\left(
\begin{array}
[c]{c}%
a^{\prime}\\
b^{\prime}\\
c^{\prime}%
\end{array}
\right)  .
\]


It is my hope and intention in writing this little book that you engage with
and enjoy this uniform way of understanding all two-dimensional geometries as
much as I did!

\begin{remark}
Special message to current or future teachers of high school geometry: Parts
II and III of this book are especially relevant to your teaching of the
subject. Look especially closely at the treatment of congruence (rigid
motion), similarity (dilation), circles, expressing geometric properties with
equations, and geometric measurement and dimension, and compare them with the
high school geometry sections of the Common Core State Standards in
Mathematics. The latter can be found at:
\end{remark}

\begin{center}
http://www.corestandards.org/Math/Content/HSG/introduction.
\end{center}

A useful companion course to one based on this book, one that might be called
\textit{Geometry for Teaching}, would explicitly make the connections between
the material covered as in this book and what you do (or will do) in your high
school geometry classroom. The idea is \textit{not} that the material we will
cover in Parts II and III will tell you how to teach that material but rather
that the treatment given here will give you the depth and breadth of geometric
understanding that will allow you to design what you teach and bring it into
your classroom in ways that those who lack that understanding cannot.

\begin{remark}
This book can also be used as a bridge to a first course in Riemannian
geometry. It treats the case of two-dimensional geometries that are
homogeneous, that is, that look the same at all their points. But to treat
these geometries efficiently, we introduce the notion of changing coordinates
for the geometry without changing the geometry itself. It is that notion that
allowed geometers to treat surfaces and higher-dimensional smooth spaces that
look different at different points, ones that can often not be treated at all
their points using a single set of coordinates.
\end{remark}
