\documentclass{ximera}

\input{../preamble.tex}

\title{Relationship between central and stereographic projection
  coordinates}

\begin{document}
\begin{abstract}
Now we study how to move between central and stereographic projection.
\end{abstract}
\maketitle


\subsection*{Two addresses for the same point in $K$-geometry}

Finally we should compare the relationship between the two kinds of
coordinates for the set%
\[
\left\{  \left(  x,y,z\right)  \in\mathbb{R}^{3}:1=K\left(  x^{2}%
+y^{2}\right)  +z^{2}\right\}
\]
that we have been exploring, namely central projection coordinates%
\begin{align*}
x_{c}  &  =\frac{x}{z}\\
y_{c}  &  =\frac{y}{z}
\end{align*}
and stereographic projection coordinates%
\begin{align*}
x_{s}  &  =\frac{2x}{z+1}\\
y_{s}  &  =\frac{2y}{z+1}.
\end{align*}
To do this we use the Problems in which we wrote $\left(  x,y,z\right)  $ in
the set
\[
\left\{ \left( x,y,z\right) \in\mathbb{R}^{3}:1=K\left( x^{2}%
+y^{2}\right) +z^{2}\right\}
\] 
as functions of $\left(x_{c},y_{c}\right)$ and
$\left(x_{s},y_{s}\right)$ respectively. Namely
\begin{align*}
x  &  =\frac{x_{c}}{\sqrt{K\left(  x_{c}^{2}+y_{c}^{2}\right)  +1}}\\
y  &  =\frac{y_{c}}{\sqrt{K\left(  x_{c}^{2}+y_{c}^{2}\right)  +1}}\\
z  &  =\frac{1}{\sqrt{K\left(  x_{c}^{2}+y_{c}^{2}\right)  +1}}%
\end{align*}
and%
\begin{align*}
x  &  =\frac{x_{s}}{1+\frac{K}{4}\left(  x_{s}^{2}+y_{s}^{2}\right)  }\\
y  &  =\frac{y_{s}}{1+\frac{K}{4}\left(  x_{s}^{2}+y_{s}^{2}\right)  }\\
z  &  =\frac{1-\frac{K}{4}\left(  x_{s}^{2}+y_{s}^{2}\right)  }{1+\frac{K}%
{4}\left(  x_{s}^{2}+y_{s}^{2}\right)  }.
\end{align*}
The rest is simple algebra. [MJG,258]

\begin{problem}
Do the algebra to write the explicit formulas for%
\[
\left(  x_{s}\left(  x_{c},y_{c}\right)  ,y_{s}\left(  x_{c},y_{c}\right)
\right)
\]
and%
\[
\left(  x_{c}\left(  x_{s},y_{s}\right)  ,y_{c}\left(  x_{s},y_{s}\right)
\right)  .
\]

\end{problem}

\subsection*{Plane sections of $K$-geometry}

\begin{problem}
Suppose we intersect $K$-geometry
\[
\left\{ \left( x,y,z\right) \in\mathbb{R}^{3}:1=K\left( x^{2}
+y^{2}\right) +z^{2}\right\}
\]
with a plane
\[
ax+by+z=0.
\]
\begin{enumerate}
\item Find the equation for the resulting path in central projection coordinates.
\item Show the equation the resulting path in stereographic projection
  coordinates is%
\[
\left(  x_{s}-\frac{2a}{K}\right)  ^{2}+\left(  y_{s}-\frac{2b}{K}\right)
^{2}=\frac{4\left(  K+a^{2}+b^{2}\right)  }{K^{2}}.
\]
\item What is the equation for the resulting path in stereographic projection
coordinates if we intersect the $K$-geometry with a plane given by%
\[
ax+by=0,
\]
that is, a plane containing the $z$-axis?
\end{enumerate}
\end{problem}

\subsection*{When $K$ is negative, there are asymptotic cones}

Notice that, if $K<0$, the equation of $K$-geometry becomes%
\[
z^{2}-\left\vert K\right\vert \left(  x^{2}+y^{2}\right)  =1.
\]
Thus $K$-geometry forms a $2$-sheeted hyperboloid with the $z$-axis as major
axis. The hyerboloid is obtained by rotating the hyperbola%
\[
z^{2}-\left\vert K\right\vert x^{2}= 1
\]
in the $\left(  x,z\right)  $-plane around the $z$-axis. We will only consider
the sheet on which $z$ is positive as forming the $K$-geometry.

Now the asymptotes of this last hyperbola are the pair of lines%
\[
\left(  z+\left\vert K\right\vert ^{1/2}x\right)  \left(  z+\left\vert
K\right\vert ^{1/2}x\right)  =z^{2}-\left\vert K\right\vert x^{2}=0.
\]
Rotating the asymptotes around the $z$-axis we obtain the \textit{asymptotic
cone}%
\[
z^{2}-\left\vert K\right\vert \left(  x^{2}+y^{2}\right)  =0
\]
for $K$-geometry.

Since the entire $K$-geometry lies inside the asymptotic cone, the central
projection coordinates $\left(  x_{c},y_{c}\right)  $ only correspond to
points in $K$-geometry when $\left(  x_{c},y_{c},1\right)  $ lies
\textit{inside} the asymptotic cone, that is, when%
\[
1-\left\vert K\right\vert \left(  x^{2}+y^{2}\right)  >0.
\]
Said otherwise, the disk of radius $\left\vert K\right\vert ^{-1/2}$ around
$\left(  0,0\right)  $ in the $\left(  x_{c},y_{c}\right)  $-plane captures
all the points $\left(  x,y,z\right)  $ of $K$-geometry under central projection.

\begin{problem}\hfil
\begin{enumerate}
\item Show that the slopes of the asymptotes to the hyperbola 
\[
z^{2}-\left\vert K\right\vert x^{2}=1 
\]
are the limits as $x$ goes to $\pm\infty$ of the slopes of the lines
through $\left( 0,0\right) $ and the point $\left( x,z\right) $ on the
hyperbola.

\item Show that the slopes of the asymptotes to the hyperbola
\[
z^{2}-\left\vert K\right\vert x^{2}=1 
\]
are the limits as $x$ goes to $\pm\infty$ of the slopes of the lines
through $\left( 0,-1\right) $ and the point $\left( x,z\right) $ on
the hyperbola.

\item Use b) to compute the radius of the disk around $\left(
  0,0\right) $ in the $\left( x_{s},y_{s}\right) $-plane that captures
  all the points $\left( x,y,z\right) $ of $K$-geometry under
  stereographic projection.
\end{enumerate}
\end{problem}


\end{document}



