\documentclass{ximera}

\input{../preamble.tex}

\title{Connections to special relativity}

\begin{document}
\begin{abstract}
Here we see a connection to special relativity.
\end{abstract}
\maketitle

In this section we will show a connection to special
relativity. Underlying all of this work is the notion of a
\textit{group action}.

\begin{definition}
  If $G$ is a group and $X$ is a set, then a (right) \dfn{group
    action} is a map (written as $x\cdot g$, for all $g\in G$ and
  $x\in X$) satisfying the following properties:
  \begin{enumerate}
  \item $(x\cdot g_1)\cdot g_2 = x\cdot (g_1\cdot g_2)$, for all $g_1,
    g_2\in G$ and $x\in X$.
  \item $x\cdot I = x$ for all $x\in X$ where $I$ is the identity
    element of the group.
  \end{enumerate}
\end{definition}
As we work, you should try to see where we are using group actions.

\section{Galilean transformations of velocity}



In this case we will let points in central projection coordinates
$(x_c,y_c)$ represent ``observed velocities.'' We have shown that
orthogonal matrices in $K$-warped space form a group. That is we have
shown that
\begin{enumerate}
\item multiplication of orthogonal matrices is associative, 
\item the product of two orthogonal matrices is orthogonal,
\item the identity matrix is orthogonal,
\item the inverse matrix $M^{-1}$ of a orthogonal matrix $M$ is orthogonal.
\end{enumerate}
The group of orthogonal $3\times 3$ matrices is denoted as
$O(3)$. We'll denote the group of $K$-orthogonal matrices as
$O_K(3)$. When $K<0$, we already know three (classes of) elements of this group:
\[
I =
\begin{bmatrix}
  1 & 0 & 0\\
  0 & 1 & 0\\
  0 & 0 & 1
\end{bmatrix},
\qquad
M_\theta=
\begin{bmatrix}
  \cos\theta & \sin\theta & 0\\
  -\sin\theta & \cos\theta & 0\\
  0 & 0 & 1
\end{bmatrix},
\]
\[
N_\psi=\begin{bmatrix}
\cosh\psi & 0 & |K|^{1/2}\cdot\sinh\psi\\
0 & 1 & 0\\
|K|^{-1/2}\cdot\sinh\psi & 0 & \cosh\psi
\end{bmatrix}
\]
If we have an observer at point $O = (0,0)$ in central projection
coordinates, we can denote any other observed velocity in the plane as
a point $(x_c,y_c)$.

\begin{problem}
  With the setting described above in mind, what does the point
  $(.3,.3)$ represent?
\end{problem}

Now if we have a configuration of points in this velocity space, we
would like to answer the following question:

\begin{quote}
  If a new observer $O'$ views the original observer $O$ as moving at
  a velocity of $(u,0)$, how do we transform all other velocities
  measured by the original observer?
\end{quote}

\begin{problem}
  Can you give a naive solution to this problem?
\end{problem}
Even if you can give a naive solution to this problem, we will think
of this problem in central projection coordinates. Eventually we want
to solve a much harder problem where the naive solution will fail,
however the solution using central projection coordinates will
prevail. 

We'll start by working when $K<0$, then we will take the limit as $K$
goes to zero, and see what we find.

\begin{problem}
  Convert the $K$-rigid motion
  \[
  N_\psi=\begin{bmatrix}
  \cosh\psi & 0 & |K|^{1/2}\cdot\sinh\psi\\
  0 & 1 & 0\\
  |K|^{-1/2}\cdot\sinh\psi & 0 & \cosh\psi
\end{bmatrix}
  \]
  to a rigid motion of the $(x_c,y_c)$-plane. Call your answer $\nu_\psi$.
  \begin{hint}
    Either write down the answer from a previous exercise or recall
    that a rigid motion in $(x,y,z)$-coordinates
    \[
    M=\begin{bmatrix}
    m_{11} & m_{12} & m_{13}\\
    m_{21} & m_{22} & m_{23}\\
    m_{31} & m_{32} & m_{33}
    \end{bmatrix}
    \]
    converts to
    \[
    \mu(x_c,y_c) = \left(
    \frac{x_c\cdot m_{11} + y_c\cdot m_{21} + m_{31}}{x_c\cdot m_{13} + y_c\cdot m_{23} + m_{33}},
    \frac{x_c\cdot m_{12} + y_c\cdot m_{22} + m_{32}}{x_c\cdot m_{13} + y_c\cdot m_{23} + m_{33}}
    \right).
    \]
  \end{hint}
  \begin{freeResponse}
    Write
    \[
    \nu_\psi(x_c,y_c) = \left(\frac{x_c\cdot\cosh\psi+|K|^{-1/2}\cdot\sinh\psi}{x_c\cdot|K|^{1/2}\cdot\sinh\psi+\cosh\psi},\frac{y_c}{x_c\cdot|K|^{1/2}\cdot\sinh\psi+\cosh\psi}\right).
    \]
  \end{freeResponse}
\end{problem}

We need to be translating velocities by a constant amount in central
projection. To do this, we must set $\psi =\mathrm{artanh}\left(u\cdot|K|^{1/2}\right)$.

\begin{problem}
  Substitute $\mathrm{artanh}\left(u\cdot|K|^{1/2}\right)$ for $\psi$ in your formula above.
  \begin{hint}
    \begin{align*}
    \cosh(\mathrm{artanh}(x)) &= \frac{1}{\sqrt{1-x^2}},\\
    \sinh(\mathrm{artanh}(x)) &= \frac{x}{\sqrt{1-x^2}}.
    \end{align*}
  \end{hint}
  \begin{freeResponse}
    Write
    \begin{align*}
      \nu_\psi(x_c,y_c) &= \left(\frac{x_c\cdot\cosh\psi+|K|^{-1/2}\cdot\sinh\psi}{x_c\cdot|K|^{1/2}\cdot\sinh\psi+\cosh\psi},\frac{y_c}{x_c\cdot|K|^{1/2}\cdot\sinh\psi+\cosh\psi}\right)\\
      &= \left(\frac{\frac{x_c}{\sqrt{1-\left(u\cdot|K|^{1/2}\right)^2}} + \frac{|K|^{-1/2}\cdot u\cdot|K|^{1/2}}{\sqrt{1-\left(u\cdot|K|^{1/2}\right)^2}}}
         {\frac{x_c\cdot |K|^{1/2}\cdot u \cdot |K|^{1/2}}{\sqrt{1-\left(u\cdot|K|^{1/2}\right)^2}}+\frac{1}{\sqrt{1-\left(u\cdot |K|^{1/2}\right)^2}}},
         \frac{y_c}{\frac{x_c\cdot |K|^{1/2}\cdot u \cdot |K|^{1/2}}{\sqrt{1-\left(u\cdot|K|^{1/2}\right)^2}}+\frac{1}{\sqrt{1-\left(u\cdot |K|^{1/2}\right)^2}}}\right)\\
         &= \left(\frac{\frac{x_c}{\sqrt{1-u^2\cdot|K|}} + \frac{u}{\sqrt{1-u^2\cdot|K|}}}
         {\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}},
         \frac{y_c}{\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}}\right).
         %&= \left(\frac{x_c + u}{x_c\cdot u \cdot |K|+1},\frac{y_c\cdot \sqrt{1-u^2\cdot|K|}}{x_c\cdot u \cdot |K|+1}\right).
    \end{align*}
  \end{freeResponse}
\end{problem}

You have just written down a transformation that will work for any
negative value of $K$. We need this to work for $K=0$.

\begin{problem}
  Take the limit as $K$ goes to zero in your answer above.
  \begin{freeResponse}
    Write
    \[
    \lim_{K\to 0} \left(\frac{\frac{x_c}{\sqrt{1-u^2\cdot|K|}} + \frac{u}{\sqrt{1-u^2\cdot|K|}}}
         {\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}},
         \frac{y_c}{\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}}\right)
         \]
         \begin{align*}
           &= \lim_{K\to 0}\left(\frac{x_c + u}{x_c\cdot u \cdot |K|+1},\frac{y_c\cdot \sqrt{1-u^2\cdot|K|}}{x_c\cdot u \cdot |K|+1}\right)\\
           &=(x_c+u,y_c).
         \end{align*}
  \end{freeResponse}
\end{problem}


At this point we are working in the \textit{euclidean plane}, that is,
central projection when $K=0$.
The answer to our question:
\begin{quote}
  If a new observer $O'$ views the original observer $O$ as moving at
  a velocity of $(u,0)$, how do we transform all other velocities
  measured by the original observer?
\end{quote}
is almost too easy, as it is that we apply the transformation
\[
\nu_u(x_c,y_c) = (x_c+u,y_c).
\]
Before you object that we have done nothing but ``made mathematics
difficult'' we preemptively retort:
\begin{center}
  \textbf{We are building a foundation for future work.}
\end{center}

\begin{problem}
  Can you rephrase what we are doing in terms of group actions?
\end{problem}




\section{Hyperbolic transformations of velocity}

Armed with experience from above, let's now suppose we are again attacking the question:
\begin{quote}
  If a new observer $O'$ views the original observer $O$ as moving at
  a velocity of $(u,0)$, how do we transform all other velocities
  measured by the original observer?
\end{quote}
however, this time we are working not on the euclidean plane, but on
the Klein disk, when $K=-1$. The beauty of our work above is that it
can be directly applied in this case:

\begin{problem}
  Show that when $u\in [0,1)$ and $K=-1$,
  \[
  \nu_u(x_c,y_c) = \left(\frac{x_c + u}{x_c\cdot u + 1},\frac{y_c\cdot \sqrt{1-u^2}}{x_c\cdot u+1}\right).
  \]
  \begin{freeResponse}
    Write
    \begin{align*}
    \nu_u(x_c,y_c)&=\left(\frac{\frac{x_c}{\sqrt{1-u^2\cdot|K|}} + \frac{u}{\sqrt{1-u^2\cdot|K|}}}
         {\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}},
         \frac{y_c}{\frac{x_c\cdot u \cdot |K|}{\sqrt{1-u^2\cdot|K|}}+\frac{1}{\sqrt{1-u^2\cdot |K|}}}\right)\\
         &=\left(\frac{x_c + u}{x_c\cdot u \cdot |K|+1},\frac{y_c\cdot \sqrt{1-u^2\cdot|K|}}{x_c\cdot u \cdot |K|+1}\right)
    \end{align*}
    and set $K=-1$ to find:
    \[
    \nu_u(x_c,y_c)=\left(\frac{x_c + u}{x_c\cdot u + 1},\frac{y_c\cdot \sqrt{1-u^2}}{x_c\cdot u+1}\right).
    \]
  \end{freeResponse}
\end{problem}

For those who are in-the-know, a minor miracle has just occurred. The transformation
\[
  \nu_u(x_c,y_c) = \left(\frac{x_c + u}{x_c\cdot u + 1},\frac{y_c\cdot \sqrt{1-u^2}}{x_c\cdot u+1}\right),
\]
is in fact the velocity transformation for Einstein's special theory
of relativity when one takes the speed of light to be $c=1$!

\begin{problem}
Show that the transformation of the Klein disk $\nu_u$ is
given by the transformation of $K$-warped space
\[
N_u = 
\begin{bmatrix}
  \frac{1}{\sqrt{1-u^2\cdot|K|}} & 0 & \frac{u\cdot|K|}{\sqrt{1-u^2\cdot|K|}}\\
  0 & 1 & 0\\
  \frac{u}{\sqrt{1-u^2\cdot|K|}} & 0 & \frac{1}{\sqrt{1-u^2\cdot|K|}}
\end{bmatrix}
\]
when $K=-1$.
\end{problem}

\begin{problem}
  Show that
  \[
  N_u = 
  \begin{bmatrix}
    \frac{1}{\sqrt{1-u^2\cdot|K|}} & 0 & \frac{u\cdot|K|}{\sqrt{1-u^2\cdot|K|}}\\
    0 & 1 & 0\\
    \frac{u}{\sqrt{1-u^2\cdot|K|}} & 0 & \frac{1}{\sqrt{1-u^2\cdot|K|}}
  \end{bmatrix}
  \]
  is a $K$-rigid motion.
\end{problem}

The upshot to all of these computations is that transformations of the form:
\[
\nu_u(x_c,y_c) = \left(\frac{x_c + u}{x_c\cdot u + 1},\frac{y_c\cdot \sqrt{1-u^2}}{x_c\cdot u+1}\right),
\]
correspond exactly to a rigid motion of the surface
\[
-(x^2+y^2) + z^2 =1
\]
with the $K$-dot product. One imaginative interpretation of these
computations is that we are in fact living on a hyperbolic surface,
and that we experience our world through the lens of a central
projection. What an amazing thought!


%% \section{Hyperbolic transformations of position}


%% THIS IS A DRAFT


%% Here we must set
%% \[
%% K = \frac{-1}{T^2}\qquad \text{or}\qquad K = \frac{-1}{H+T^2}.
%% \]
%% Where $T$ is another time parameter different from $t$.
%% Computationally $T=t$ and $K= \frac{-1}{t^2}$ seems to work out, but
%% it makes very little physical sense to me what this means, or why this
%% should be the case.

%% Write
%% \[
%% N_{u\cdot t} = 
%% \begin{bmatrix}
%%   \frac{1}{\sqrt{1-u^2\cdot t^2\cdot|K|}} & 0 & \frac{u\cdot t^2\cdot|K|}{\sqrt{1-u^2\cdot t^2\cdot|K|}}\\
%%   0 & 1 & 0\\
%%   \frac{u\cdot t^2}{\sqrt{1-u^2\cdot t^2\cdot|K|}} & 0 & \frac{1}{\sqrt{1-u^2\cdot t^2\cdot|K|}}
%% \end{bmatrix}
%% \]
%% If $K=\frac{-1}{t^2}$, then 
%% \[
%% N_{u\cdot t} = 
%% \begin{bmatrix}
%%   \frac{1}{\sqrt{1-u^2}} & 0 & \frac{u}{t\cdot\sqrt{1-u^2}}\\
%%   0 & 1 & 0\\
%%   \frac{u\cdot t}{\sqrt{1-u^2}} & 0 & \frac{1}{\sqrt{1-u^2}}
%% \end{bmatrix}
%% \]
%% and this is a $K$-rigid motion in this case. Moreover, here
%% \[
%% \nu_{u\cdot t}(v\cdot t,0) = \left(\frac{(v+u)\cdot t}{\sqrt{1+v\cdot u}},0\right)  
%% \]









\begin{problem}
Summarize the results from this section. In particular, indicate which
results follow from the others.
\begin{freeResponse}
\end{freeResponse}
\end{problem}




\end{document}
