\documentclass[newpage,hints,handout]{ximera}

\input{../preamble.tex}

\title{Surface area and volume of the $R$-sphere}
\begin{document}
\begin{abstract}
In this activity we begin to explore spherical geometry.
\end{abstract}
\maketitle


We are now going to study the geometry of the $R$-sphere in Euclidean
$3$-space. This is the sphere of radius $R$ in normal $3$-space.  We
will show why there is a factor of $1/3$ in many formulas for volumes
in $3$-dimensional Euclidean geometry, just like there is a factor of
$1/2$ in many formulas for areas in $2$-dimensional Euclidean
geometry.



\section{Circumference and area of circles}
\begin{listOutcomes}
 \item Prove that the circumference of a circle with radius $1$ is $2\pi$.
 \item Give two different proofs that the area of a circle with radius $1$ is $pi$.
\end{listOutcomes}

We begin with perhaps the most basic fact of all about circles in
Euclidean geometry. We will use calculus to compute the perimeter of a unit
circle.

%%%%%%FOR ONLINE, TURNED THE DRAWING INTO THE SET UP FOR THE NEXT PROBLEM%%%%%%
%\begin{problem}
{\bf Problem Set Up} Consider a unit circle with $n$ equal inscribed triangles meeting at the center.
Here we have shown the $n$th triangle:
\begin{image}
\begin{tikzpicture}[geometryDiagrams]
\draw [domain=-20:65] plot ({4*cos(\x)}, {4*sin(\x)});
\coordinate (A) at (4,0);
\coordinate (O) at (0,0);
\coordinate (C) at (2.83,2.83);
\draw[thin] (A)--(O)--(C)--(A);
\node[left] at (3.41,1.41) {$b_n$};
\node[right] at (4,0) {$(1,0)$};
\node[left] at (0,0) {$(0,0)$};
%\node[above] at (C) {$P$};
\tkzMarkAngle[size=0.7cm,thin](A,O,C)
\tkzLabelAngle[pos = 0.5](A,O,C){$\theta_n$}
\end{tikzpicture}
\end{image}

Find $\theta_n$ and $b_n$ in terms of $n$.
%\end{problem}


\begin{problem} Write $\theta_n$ and $b_n$ in terms of $n$.

Explain what the limit
\[
\lim_{n\to\infty} \left(n\cdot b_n\right)
\]
computes and compute the limit.
\end{problem}



\begin{problem}
The circle of radius $1$ has circumference $2\pi$. Use this to show
that the circle of radius $1$ has area $\pi$.
\begin{image}
\begin{tikzpicture}[geometryDiagrams]
\draw (0,0) circle (2cm);
\foreach \coeff in {0,10,20,...,350}
         {
           \draw[thin] (0,0)--({2*cos(\coeff)},{2*sin(\coeff)});
         }
\end{tikzpicture}
\end{image}

\begin{hint}
Approximate a rectangle by rearranging the slices in the picture.
Compute the area of the ``rectangle.''
\end{hint}
\begin{freeResponse}
Slice up the circle as indicated and rearrange into the following
``rectangle.'' 
\begin{image}
\begin{tikzpicture}[geometryDiagrams] % r= 2cm
  \foreach \coeff in {0,1,2,...,17}
   {
     \draw[thin] ({2*\coeff *sin(5)},0)--({(2*\coeff+ 2)* sin(5)},{2*cos(5)});
     \draw[thin] ({(2*\coeff + 2)*sin(5)},{2*cos(5)})--({(2+2*\coeff)*sin(5)},0); %one too many!
   }
   \draw[thin] ({2*18 *sin(5)},0)--({(2*18+ 2)* sin(5)},{2*cos(5)});
   \draw (0,0)--({2*18 *sin(5)},0);
   \draw ({2*sin(5)},2)--({2*19 *sin(5)},2);
\end{tikzpicture}
\end{image}
Note, while this is not truly a rectangle, as the slices become finer,
and finer, the shape will more, and more, closely approximate a
rectangle. This rectangle has perimeter $2+2\pi$, and hence is a
$1\times \pi$ rectangle. Thus the area of the rectangle is $\pi$ as is
the area of the initial circle.
\end{freeResponse}
\end{problem}


Not content to give just one explanation, let's give another. The
techniques we use in this explanation will be used in later problems.

\begin{problem}
Again, consider the circle of radius $1$ with circumference $2\pi$.
Suppose we inscribe $n$ triangles (with one vertex at the center of
the circle), here we have shown the $n$th triangle:
\begin{image}
\begin{tikzpicture}[geometryDiagrams]
\draw [domain=-20:65] plot ({4*cos(\x)}, {4*sin(\x)});
\coordinate (A) at (4,0);
\coordinate (B) at (3.41,1.41);
\coordinate (O) at (0,0);
\coordinate (C) at (2.83,2.83);
\draw[thin] (A)--(O)--(C)--(A);
\draw[thin] (B)--(O);
\node at (2,1.1) {$h_n$};
\node[below] at (3.1,2) {$b_n$};

\tkzMarkRightAngle[thin](A,B,O)

\end{tikzpicture}
\end{image}
In each triangle, label $b_n$ and $h_n$, where $b_n$ is conceptualized
as a ``base'' and $h_n$ is a ``height.'' Find
\[
\lim_{n\to \infty} h_n
\]
and explain your reasoning.
\begin{freeResponse}
As $n$ goes to infinity, $h_n$ goes to the radius length of the radius
of the circle. Hence
\[
\lim_{n\to \infty} h_n = 1.
\]
\end{freeResponse}
\end{problem}


\begin{problem}
Consider the limit
\[
\lim_{n\to\infty} \frac{n\cdot b_n\cdot h_n}{2}.
\]
Explain what this limit is computing, and compute this limit.
\begin{freeResponse}
This limit is computing the sum of the areas of the triangles as $n$
goes to infinity. This will be the area of the circle of radius
$1$. From using substitution from our work above, we see
\begin{align*}
\lim_{n\to\infty} \frac{n\cdot b_n\cdot h_n}{2} &= \frac{1\cdot 2\pi}{2}\\
&=\pi.
\end{align*}
\end{freeResponse}
\end{problem}


\section{Surface area of spheres}
\begin{listOutcomes}
\item Prove the formula for the surface area of a sphere.
\end{listOutcomes}

To compute the surface area of the sphere of radius $R$ in $3$-dimensional
Euclidean space, we will show that its surface area is equal to the surface
area of something we can lay out flat. The argument for this goes way back to
the great physicist and mathematician, Archimedes of Alexandria, in the $2$nd
century B.C. To follow his argument, we have to begin by computing the area of
a `lamp shade' or `collar.' We think of a circular collar as in the figure
below%
\begin{image}
\begin{tikzpicture}
    \draw (-1,0) arc (180:360:1cm and 0.5cm);
    \draw (-1,0) arc (180:0:1cm and 0.5cm);
    \draw (-2,-3) arc (180:370:2cm and 1cm);
    \draw[dashed] (-2,-3) arc (180:10:2cm and 1cm);
    \draw(-2,-2.9)  -- (-1,0);
    \draw(2,-2.9)   -- (1,0);
    \node [left] at (2,-1) {$s$};
    \shade[left color=blue!5!white,right color=blue!60!white,opacity=0.3] (-1,0) arc (180:360:1cm and 0.5cm) -- (2,-3) arc (360:180:2cm and 1cm) -- cycle;
    \shade[left color=blue!5!white,right color=blue!60!white,opacity=0.3] (0,0) circle (1cm and 0.5cm);
\end{tikzpicture}
\end{image}
as approximated by an arrangement of trapezoids. To achieve this, we
approximate the bottom circle of the collar by an inscribed regular $n$-gon
whose vertices are the points of intersection with the slant lines in the
figure. Similarly approximate the top circle by an inscribed regular $n$-gon
positioned directly above the bottom one, again with vertices given by the
points of intersection with the slant lines. Complete a side of the bottom
$n$-gon and the side of the top $n$-gon directly above it to a trapezoid by
adjoining the two slant lines in the figure that connect endpoints. Let
\begin{itemize}
\item $b_{n}$ denote the length of a side of the bottom regular
  $n$-gon and let 
\item $t_{n}$ denote the length of a side of the top $n$-gon. 
\end{itemize}
Then the trapezoid has area
\[
\left(  \frac{b_{n}+t_{n}}{2}\right)  \cdot h_{n}%
\]
where $h_{n}$ is the vertical height of the trapezoid. The collar is
approximated by the union of these $n$ trapezoids, so the area of the collar
is approximated by the sum of the areas of the $n$ congruent trapezoids,
namely%
\[
n\cdot \left(  \frac{b_{n}+t_{n}}{2}\right)
\cdot h_{n}=\left(  \frac{n\cdot %
b_{n}+n\cdot t_{n}}{2}\right) \cdot h_{n}.
\]
As $n$ goes to infinity, the area of the approximation approaches the area of
the collar. But if 
\begin{itemize}
\item $c_{b}$ is the circumference of the bottom circle and 
\item $c_{t}$ is the circumference of the top circle and 
\item $s$ is the slant height of the collar as shown in the above
  figure,
\end{itemize}
then
\begin{align*}
\lim_{n\to \infty} n\cdot b_{n}  &  =c_{b}\\
\lim_{n\to \infty} n \cdot t_{n}  &  =c_{t}\\
\lim_{n\to \infty} h_{n}  &  =s.
\end{align*}


\begin{problem}
Let
\begin{itemize}
\item $r_b$ be the radius of the circle defining the base of the collar,
\item $r_t$ be the radius of the circle defining the top of the collar,
\item $r_a$ be the average of $r_b$ and $r_t$,
\item $s$ be the slant height of the collar.
\end{itemize}
Explain why the area of the collar is%
\[
\pi \cdot \left( r_{b}+r_{t}\right)\cdot s.
\]
(This can also be written as $2\pi\cdot r_{a}\cdot s$.)
\end{problem}


\begin{theorem}
 The surface area of the sphere of radius $R$ is the same as the
surface area of the label of the smallest can into which the sphere will fit.%
\pgfdeclareradialshading{ballshading}{\pgfpoint{-10bp}{10bp}}
 {color(0bp)=(gray!40!white); 
 color(9bp)=(gray!75!white);
 color(18bp)=(gray!70!blue); 
 color(25bp)=(gray!50!blue); 
 color(50bp)=(black)}
\begin{image}
\begin{pgfpicture}
\pgfellipse[]{\pgfxy(0,-2)}{\pgfxy(0,.7)}{\pgfxy(2,0)}
\pgfstroke
  \pgfpathcircle{\pgfpoint{0cm}{0cm}}{2cm}
  \pgfshadepath{ballshading}{20}
  \pgfusepath{}
 \pgfxyline(2,2)(2,-2)
 \pgfxyline(-2,2)(-2,-2)
\pgfellipse[]{\pgfxy(0,2)}{\pgfxy(0,.7)}{\pgfxy(2,0)}
\pgfstroke
\end{pgfpicture}
\end{image}
Namely the surface area of the sphere of radius $R$ is
\[
2\pi R\cdot 2R=4\pi R^{2}.
\]

\end{theorem}

\begin{problem}
Show why the above theorem is true.

\begin{hint}
Slice the picture above into $n$ horizontal slices. Approximate the
piece of the surface of the sphere between the $i$th pair of
successive slices by a collar $C_{i}$. Let $a\left( C_{i}\right) $
denote the area of $C_{i}$, let $r_{i}$ denote its average radius and
Explain why the surface area of the sphere is
\[
\lim_{n\to \infty}\sum_{i=1}^{n} a(C_i).
\]
Explain further how we can conclude that the area of the sphere is
given by
\[%
\lim_{n\to \infty}\sum_{i=1}^{n} 2\pi\cdot r_{i}\cdot s_{i}.
\]

\end{hint}
\begin{hint}
Let $h_{i}$ denote the vertical height of the label on the can between
the $i$th pair of successive slices. Explain why the area of the label is exactly%
\[%
\sum_{i=1}^{n} 2\pi R \cdot h_{i}.
\]
\end{hint}
\begin{hint}
Explain why the relationship between each $r_{i}$, $s_{i}$ and $h_{i}$
is given by the picture below.
\begin{image}
\begin{tikzpicture}[geometryDiagrams]
\coordinate (A) at (0,1);
\coordinate (B) at (4,1);
\coordinate (C) at (4,-1);
\coordinate (D) at (.5,2);
\coordinate (E) at (-.5,0);
\coordinate (F) at (-.5,2);
\draw (A)--(B)--(C)--cycle;
\draw (D)--(E)--(F)--cycle;
\node[above] at (2,1) {$r_i$};
\node[above] at (2,0) {$R$};
\node[left] at (-.5,1) {$h_i$};
\node[left,above] at (-.2,1) {$s_i$};
\end{tikzpicture}
\end{image}
Now use facts about similar triangles to explain why
\[
r_{i}\cdot s_{i}=h_{i}\cdot R.
\]
\end{hint}
\end{problem}




\section{Volumes of pyramids and spheres}
\begin{listOutcomes}
\item State the magnification principle.
 \item Prove the formula for the volume of a rectangular base pyramid.
\end{listOutcomes}

We will start by looking at the volumes of pyramids.

\begin{problem} Show that an $r\times r\times r$ cube can be constructed
from three equal pyramids with an $r\times r$ square base. Conclude
that the volume of each pyramid is $1/3$ the volume of the cube,
namely
\[
\frac{r^{3}}{3}.
\]

\begin{hint}
Suppose the cube had a hollow interior and infinitely thin
faces. Put your (infinitely tiny) eye at one vertex of the cube and
look inside. How many faces of the cube can you see?
\end{hint}

\begin{freeResponse}
\begin{image}
\begin{tikzpicture}
\coordinate (A) at (0,0,0);
\coordinate (B) at (4,0,0);
\coordinate (C) at (4,4,0);
\coordinate (D) at (0,4,0);
\coordinate (A') at (0,0,4);
\coordinate (B') at (4,0,4);
\coordinate (C') at (4,4,4);
\coordinate (D') at (0,4,4);



\draw[fill=red,opacity=0.3] (A)--(B)--(C)--(D)--cycle; % base
\draw[fill=blue,opacity=0.3] (A')--(B')--(C')--(D')--cycle; % base
%% \draw[fill=orange,opacity=0.3] (B1)--(B2)--(peak);
%% \draw[fill=yellow,opacity=0.3] (B2)--(B3)--(peak);
%% \draw[fill=green,opacity=0.3] (B3)--(B4)--(peak);
%% \draw[fill=blue,opacity=0.3] (B4)--(B5)--(peak);
%% \draw[fill=purple,opacity=0.3] (B5)--(B1)--(peak);


\draw (A)--(B)--(C)--(D)--cycle;
\draw (A)--(A');
\draw (B)--(B');
\draw (C)--(C');
\draw (D)--(D');
\draw (A')--(B')--(C')--(D')--cycle;

\draw[ultra thick] (A)--(A')--(B')--(B)--cycle;
\draw[ultra thick] (A')--(D);
\draw[ultra thick] (B)--(D);
\draw[ultra thick, red] (B')--(D);
\draw[ultra thick, red] (A)--(D);
\end{tikzpicture}
\end{image}
\end{freeResponse}
\end{problem}

We next want to show why all pyramids with an $r\times r$ square base
and vertical altitude $r$ have the same volume. That is, if we put the
vertex of the pyramid anywhere in a plane parallel to the base and at
distance $r$, the volume is unchanged.

This fact is an example of \textit{Cavalieri's Principle}: Shearing a
figure parallel to a fixed direction does not change the
$n$-dimensional measure of an object in Euclidean $n$-space. Think of
a stack of (very thin) books. We'll give a proof in Euclidean $3$-space,
and we will use the coordinates $(\hat{x},\hat{y},\hat{z})$ to denote
this space. 






\begin{problem}
Show that Cavalieri's Principle is true for the pyramid using
multivariable calculus.

\begin{hint}
Put the base of the pyramid $P$ so that its vertices are $\left(
0,0\right) $, $\left( r,0\right) $, $\left( 0,r\right) $ and $\left(
r,r\right) $ in $3$-dimensional Euclidean space. Consider the
transformation%
\[
T(x,y,z)=
\begin{bmatrix}
1 & 0 & a\\
0 & 1 & b\\
0 & 0 & 1
\end{bmatrix}
\begin{bmatrix} x \\ y \\ z\end{bmatrix}
\]
and use the 3-dimensional change of variables formula,
$$\iiint_{T(P)} f(x,y,z)dxdydz=\iiint_P f(T(x,y,z))\lvert\det DT(x,y,z)\rvert dxdydz,$$
where
$$DT(x,y,z)=\begin{bmatrix}
\frac{\partial T}{\partial x} & \frac{\partial T}{\partial y} & \frac{\partial T}{\partial z}
\end{bmatrix}.$$
(Each partial derivative is itself a vector since $T(x,y,z)$ has three
components.)
\end{hint}



\end{problem}

%\section{The magnification principle}

\textbf{The magnification principle:} If an object in Euclidean $n$-space is
magnified by factors of $r_{1},\ldots,r_{n}$, its $n$-dimensional
measure is multiplied by $r_{1}\cdots r_{n}$.
\begin{remark}
 When $n=2$ and $r_1=r_2$, this is a dilation, as in problem 31.
\end{remark}

\begin{theorem}[The magnification principle]
We now prove the magnification principle using multivariable calculus.
\end{theorem}

\begin{proof}
In Chapter 4, we used the 2-dimensional change of variables formula. In Cavalieri's Principle, we used the three dimensional change of variables formula. Now, we extend that idea to the $n$-dimensional change of variables formula:
\begin{equation}\label{changevar}
\begin{split}
 &\int\int\dots\int_{J(R)} f(x_1,x_2,\dots,x_n)dx_1 dx_2\dots dx_n\\
 &= \int\int\dots\int_{R} f(J(x_1,x_2,\dots,x_n))|\det J(x_1,x_2,\dots,x_n)|dx_1 dx_2\dots dx_n.
\end{split}\end{equation}

As we saw in the two dimensional version, when computing area, $f(x,y)=1$. Here, we are computing the $n$-dimensional version of volume, so $f(x_1,x_2,\dots,x_n)$. 

Now we translate the magnification principle into a function and into a matrix. We want to magnify $x_1$ by $r_1$, $x_2$ by $r_2$, all the way through the list to multiplying $x_n$ by $r_n$. As a function, the magnification is \[J(x_1,x_2,\dots,x_n)=(r_1 x_1,r_2 x_2,\dots, r_n x_n).\] As a matrix, $\vec x=
\left[\begin{smallmatrix}
 x_1\\x_2\\\vdots\\x_n
\end{smallmatrix}\right]$, and \[
J(\vec x)=
\begin{bmatrix}
r_{1} & 0   & \dots & 0\\
0     & r_2 & \dots & 0\\
\vdots &\vdots &\ddots & \vdots \\
0     & 0   & \dots & r_{n}%
\end{bmatrix}\vec x.
\]

Since $J(\vec x)$ is a diagonal matrix, the absolute value of the determinant is $|r_1 r_2 r_3\cdots r_n|$. Now we can plug all of our formulas into the change of variables formula \eqref{changevar}:
\begin{equation}
\begin{split}
 \int\int\dots\int_{J(R)}1dx_1 dx_2\dots dx_n
 &= \int\int\dots\int_{R}1|r_1 r_2 r_3\cdots r_n|dx_1 dx_2\dots dx_n\\
 &=|r_1 r_2 r_3\cdots r_n|\int\int\dots\int_{R}dx_1 dx_2\dots dx_n.
\end{split}\end{equation}
\end{proof}

\begin{problem}
 Explain how the statement and proof of the magnification principle is related to the statement and proof of Problem 31.
\end{problem}

\begin{problem}
Use this magnification principle to justify the volume formula%
\[
(1/3)B\cdot h
\]
for any pyramid with rectangular base of area $B$ and vertical altitude $h$.
\begin{hint}
 Use a previous problem that found the formula for the volume of a pyramid with square base of area $r^2$ and vertical altitude $r$.
\end{hint}
\end{problem}




%Now suppose we have any pyramid%
%\begin{image}
%\begin{tikzpicture}
%%\draw[thick,-stealth] (0,0,0)--(0,0,6); 
%%\draw[thick,-stealth] (0,0,0)--(0,6,0);
%%\draw[-stealth] (0,0,0)--(6,0,0);
%\coordinate (B1) at (0,0,0);
%\coordinate (B2) at (0,0,2);
%\coordinate (B3) at (2,0,1);
%\coordinate (B4) at (4,0,2);
%\coordinate (B5) at (3,0,0);
%\coordinate (peak) at (1,3,1);
%
%\draw[fill=red,opacity=0.3] (B1)--(B2)--(B3)--(B4)--(B5)--cycle; % base
%\draw[fill=orange,opacity=0.3] (B1)--(B2)--(peak);
%\draw[fill=yellow,opacity=0.3] (B2)--(B3)--(peak);
%\draw[fill=green,opacity=0.3] (B3)--(B4)--(peak);
%\draw[fill=blue,opacity=0.3] (B4)--(B5)--(peak);
%\draw[fill=purple,opacity=0.3] (B5)--(B1)--(peak);
%
%\draw[ultra thick] (B2)--(B3)--(B4)--(B5)--(peak)--(B4)--(B3)--(peak)--(B2);
%\end{tikzpicture}
%\end{image}
%with any shaped base of area $B$ and any vertical altitude $h$. Approximate
%the base as close as you want (i.e.\ $\varepsilon$-close) by tiling its interior
%with rectangles. Let $t$ denote the sum of the areas of these rectangles.
%Approximate the base as close as you want (i.e.\ $\varepsilon$-close) by
%covering it entirely with rectangles. Let $T$ denote the sum of the areas of
%these rectangles. Why is the area $B$ of the base of the pyramid caught
%between $B-\varepsilon$ and $B+\varepsilon$?
%
%\begin{problem}
%Show that the volume $V$ of the pyramid is caught between $\left(
%1/3\right) \cdot t\cdot h$ and $\left( 1/3\right) \cdot T\cdot
%h$. Argue that, given any positive real number $\varepsilon$, however
%small, the volume $V$ of the pyramid is caught between $\left(
%1/3\right) \cdot \left( B-\varepsilon\right)\cdot h$ and $\left(
%1/3\right) \cdot \left( B+\varepsilon\right) \cdot h$. Finally, show that
%\[
%V=\left(  1/3\right)  \cdot B\cdot h.
%\]
%\end{problem}

\section{Relation between volume and surface area of a sphere}
\begin{listOutcomes}
 \item Use pyramids to approximate the volume and surface area of a sphere.
 \item Use pyramids to find a relationship between the volume and surface area of a sphere.
\end{listOutcomes}
Think of a disco-ball. Its surface is approximately a sphere, but that
surface is made up of tiny flat mirrors.

\begin{problem}\hfil
\begin{enumerate}
\item Explain why you can think of the disco-ball as being made up of
  pyramids, with each pyramid having base one of the tiny mirrors and
  vertex at the interior point $O$ at the center of the disco-ball.
\item Argue that the volume of the disco-ball is $\left( 1/3\right) $
  times the distance $h$ from a mirror to $O$ times the sum of the
  areas of all the mirrors.
\end{enumerate}
\end{problem}

\begin{problem}
Argue that, as the mirrors are made to be smaller and smaller,
\begin{enumerate}
\item the sum of the areas of the mirrors approaches the surface area of a sphere,
\item the distance $h$ approaches the radius $R$ of that sphere,
\item the volume of the disco-ball approaches the volume of the sphere.
\end{enumerate}
Conclude that, for a sphere of radius $R$ in Euclidean $3$-space, the relation
between the volume $V$ of the sphere and the surface area $S$ of the sphere is
given by the formula%
\[
V=\frac{R\cdot S}{3}.
\]
\end{problem}

\begin{problem}
Use the work above to deduce a formula for the volume of a sphere.
\end{problem}




\begin{problem}
Summarize the results from this section. In particular, state the conclusion of each problem in your own words and indicate which
results follow from the others.
\begin{freeResponse}
\end{freeResponse}
\end{problem}


\end{document}
